---
title: "604 Final Project"
author: "Susan Ubogu"
---

## Final Project Purpose

The goal of this final project is to explore how different factors affect a students performance in three testing subjects: math, reading, and writing. This will be presented in a storyboard dashboard aimed at a general audience.

## Project Summary



## Understanding the Dataset

This project is based on a dataset that has information on student performance in Exams. It has the following columns:

-   gender: the gender of the student (Male/Female)

-   race/ethnicity: ethnic group (A, B, C, D, E)

-   parental level of education: the highest level of education completed by the student's parents (High School, Associate's Degree, Bachelor's Degree, Master's Degree)

-   lunch: the type of lunch a student receives (Standard, Free, Reduced)

-   test preparation course: if the student has completed a test preparation course (None/Completed)

-   math score: math exam score (0-100)

-   reading score: reading exam score (0-100)

-   writing score: writing exam score (0-100)

## Exploratory Data Analysis

Here we're going to perform a basic analysis to be able to understand the data and identify key trends.

### Descriptive Statistics

First we load the dataset into R and then we take a peek at the first 6 rows, just to make sure that it was all loaded in properly.

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrplot)
```

```{r}
data <- read.csv('/Users/susanubogu/Desktop/university assignments/fall 24/dacss 604/final project/StudentsPerformance.csv')

head(data)
dim(data)
```

Based on the data card we know there are no missing values, but we can double check that here. Since, the sum of the existing NAs is 0, we can conlude that there are none in the dataset.

```{r}
sum(is.na(data))
```

We then want to take a look at the mean, median, and standard deviation for each subject score.

```{r}
calc_stats <- function(column) {
  stats <- list(
    range = range(column, na.rm = TRUE),
    average = mean(column, na.rm = TRUE),
    std = sd(column, na.rm = TRUE),
    na = sum(is.na(column)),
    unique = length(unique(column))
  )

  return(stats)
}

print("Descriptive Statistics for Math:")
calc_stats(data$math.score)

print("Descriptive Statistics for Reading:")
calc_stats(data$reading.score)

print("Descriptive Statistics for Writing:")
calc_stats(data$writing.score)
```

### Visualizations

We then take a look at what the data looks like. Taking a look at the histogram we are able to see how the scores are spread out.

```{r}
subjects <- c("math.score", "reading.score", "writing.score")

par(mfrow = c(1, 3))
for (i in subjects) {
  hist(data[[i]], main = paste("Distribution of", gsub("\\.", " ", i)),
       xlab = "Score", col = "light blue", breaks = 20)
}
```

We then want to use box plots to see how different categorical variables affect a student's score.

```{r}
data_long <- data |>
  pivot_longer(cols = c(math.score, reading.score, writing.score),
               names_to = "Subject", values_to = "Score")

ggplot(data_long, aes(x = Subject, y = Score, fill = gender)) +
  geom_boxplot() +
  labs(title = "Score Distribution by Gender", x = "Subject", y = "Score") +
  theme_minimal()
```

```{r}
ggplot(data_long, aes(x = Subject, y = Score, fill = lunch)) +
  geom_boxplot() +
  labs(title = "Score Distribution by Lunch Type", x = "Subject", y = "Score") +
  theme_minimal()
```

Next, we want to use a bar graph to see how the average of scores was affected by test preparation and by the level of parental education.

```{r}
avg_scores <- data %>%
  group_by(test.preparation.course) %>%
  summarise(across(c(math.score, reading.score, writing.score), mean)) %>%
  pivot_longer(cols = c(math.score, reading.score, writing.score), 
               names_to = "Subject", values_to = "Average.Score")

ggplot(avg_scores, aes(x = Subject, y = Average.Score, fill = test.preparation.course)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Scores by Test Preparation Course", x = "Subject", y = "Average Score") +
  theme_minimal()
```

```{r}
avg_scores_parent_edu <- data %>%
  group_by(parental.level.of.education) %>%
  summarise(across(c(math.score, reading.score, writing.score), mean)) |>
  pivot_longer(cols = c(math.score, reading.score, writing.score),
               names_to = "Subject", values_to = "Average.Score")

ggplot(avg_scores_parent_edu, aes(x = Average.Score, y = parental.level.of.education, fill = Subject)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Scores by Parental Education Level", x = "Average Score", y = "Parental Level of Education") +
  theme_minimal()
```

We now want to use a heatmap to see if the scores are correlated to one another.

```{r}
subjects <- c("math.score", "reading.score", "writing.score")
corr_matrix <- cor(data[subjects])

corrplot(corr_matrix, method = "color", type = "upper", tl.col = "black", tl.srt = 45,
         col = colorRampPalette(c("pink", "white", "light blue"))(200))
```

### Key Findings

Based on the analysis above and the visualizations I am choosing to focus on the following findings:

-   ***Effect of Test Preparation:*** We can see that the students who had completed some test preparation did much better on the exams than students who had completed none. With completing the test prep there was at least a 5% increase in the average score, going all the way up to a 10% increase.

-   ***Gender Differences:*** We can see that the students who ideintifies as women had a higher average score in the math section. However, in the reading and writing exams the men outshone women. It is interesting to notice that the women had heavy outliers in all exams whereas we don't really see that trend with the men.

-   ***Lunch Type:*** We can see that students who had a standard lunch tended to score higher than those who were on either a free or a standard lunch.

-   ***Parental Education:*** We can see that students with parents who have completed a Master's degree had the highest scores in all exams. The general trend with this is that the more education a student's parent has gone through, the higher they tend to score on exams.

### Interactive Dashboard

If you would like to play around with the data and see how different variables interact with the average of the scores you can use this dashboard.

<a href="https://app.powerbi.com/reportEmbed?reportId=d1fb6f43-3d23-4fac-9370-ce48848e7b56&autoAuth=true&ctid=7bd08b0b-3395-4dc1-94bb-d0b2e56a497f" target="_blank">Explore the Dashboard</a>


## Models

In this section, we develop a linear regression model to predict student performance in math, reading, and writing scores based on various factors.

```{r}
lm_math <- lm(math.score ~ gender + race.ethnicity + parental.level.of.education + 
                lunch + test.preparation.course, data = data)
summary(lm_math)

lm_reading <- lm(reading.score ~ gender + race.ethnicity + parental.level.of.education + 
                   lunch + test.preparation.course, data = data)
summary(lm_reading)

lm_writing <- lm(writing.score ~ gender + race.ethnicity + parental.level.of.education + 
                   lunch + test.preparation.course, data = data)
summary(lm_writing)
```
